Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-08-16 paper/ifo-singleTone.tex
some general text editing and addressing comments
small tidying in the text

% the webcam method
% how to analyse video, python OpenCV2

Continuous wave searches look for signals which are close to monochromatic. 
In this section we consider a simple sinusoidal tone; a time at a single constant frequency. 
As described in Section~\ref{sec:ifo}, the audio is played through a speaker fixed to the back of one of the mirrors.

@@

%The resulting data is a video capture of the whole interference pattern. 
%As only the frequency of the interference ring motion is needed, w
We take data from a single off-centre pixel indicated by the red dot in Fig.~\ref{fig:interference_pattern}. 
The green-channel of the video is used as an approximation to the total intensity as the laser produces monochromatic green light (at $532\,{\rm nm}$).
As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording.
We measure a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. 
Two harmonics can also be seen in Fig.~\ref{fig:webcam_spectrum} at integer harmonics of the main peak. 
The $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks have amplitudes of $0.103$ and $0.098$ respectively as a fraction of the height of the main peak. 
They are probably due to the nonlinear system response discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}.

%\han{[to do - rewrite once James has the new numbers (also Fig 3 caption)] 
%We measure a peak amplitude at $(2.099\pm 0.008)\,{\rm Hz}$ (where the error represents the $1\sigma$ uncertainty).
%The demonstration is able to recover a single frequency tone to within two standard deviations. }
%\jam{[For the single tone result, the FWHM is 0.033 Hz. This puts the injected 2.09 Hz well within the uncertainty in the measured value of 2.099 Hz.]}
%We played this time series of intensity over time as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
%The two tones played together sound similar and don’t produce any noticeable beating.
%To quantify the precision, we recorded a signal from an injected tone of $2.09\,{\rm Hz}$. 
%From the Fourier transform shown in Figure~\ref{fig:webcam_spectrum}, we measured a peak amplitude at $(2.099\pm 0.008)\,{\rm Hz}$, a result within two standard deviations. 
%The experiment is able to recover a single tone confidently. 
%This showed that the experiment observes a single tone confidently.

%\jam{[re: quantifying the non-linear response, the first two harmonics (at 4.19 and 6.28 Hz) are 0.103 and 0.098 the height of the main peak respectively.]}

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Injected tone at a constant frequency: Fourier amplitude of the intensity time series recorded by the webcam. 
%Single injected tone results. 
%The Fourier amplitude of the intensity timeseres recorded by webcam. 
The frequency of the injection is $2.09\,{\rm Hz}$. 
%The recovered frequency peak is at $(2.099\pm 0.008)\,{\rm Hz}$. 
The recovered signal peaks at $2.099\,{\rm Hz}$ with a FWHM of $0.033\,{\rm Hz}$. 
There are also two visible harmonic peaks at integer multiples of the main peak. 
Their amplitudes as a fraction of the height of the main peak are  $0.103$ and $0.098$ for the $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks respectively. 
%\han{[to do: add new numbers from James]}
%\jam{[For the single tone result, the FWHM is 0.033 Hz. This puts the injected 2.09 Hz well within the uncertainty in the measured value of 2.099 Hz.]}
%The peak (with amplitude $17.5$) has a frequency was measured at $(2.099\pm 0.008) \mathrm{Hz}$. The actual maximum is achieved at $0\,{\rm Hz}$, but corresponds to the signal having a non-zero average value, so we don’t display or consider it.
}

\begin{comment}
As described in Section~\ref{sec:introduction}, continuous gravitational wave searches look for near monochromatic signals. 
The audio analogue of this is a simple sinusoidal tone; a constant note.

The intensity of the interference pattern is measured at a single point on the screen using a commercial USB webcam placed to view the screen (see Figure~\ref{fig:ifo_schmatic_webcam}). 
The webcam's sample rate is $30\,{\rm Hz}$, which limits the spectral content of observable signals to below 15Hz, the Nyquist frequency.
A tone under $15\,{\rm Hz}$ is played through the speaker for a mintue and the interference pattern recorded with the webcam.

The data is taken from a single off-centre pixelm, indicated by the red dot in Figure~\ref{fig:interference_pattern}. 
We take the green-channel of the video as an approximation to the total intensity as the laser is monochromatic green light (at $532\,{\rm nm}$).

We played this time series of intensity over time as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. The two tones played together sound similar and don’t produce any noticeable beating.
To quantify the precision, we recorded a signal from an injected tone of 2.09Hz. From the Fourier transform shown in Figure~\ref{fig:webcam_spectrum}, we measured a peak amplitude at $(2.099\pm 0.008)\mathrm{Hz}$, a result within two standard deviations. This showed that the experiment observes a single tone confidently.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{Fourier amplitudes of intensity time series of interference pattern as captured by a webcam for an injected tone of 2.09Hz, the amplitude 17.5 peak was measured at $(2.099\pm 0.008) \mathrm{Hz}$. The actual maximum is achieved at 0Hz, but corresponds to the signal having a non-zero average value, so we don’t display or consider it.}
	\label{fig:webcam_spectrum}
\end{figure*}
\end{comment}

>>>>

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Injected tone at a constant frequency: Fourier amplitude of the intensity time series recorded by the webcam. 
The frequency of the injection is $2.09\,{\rm Hz}$. 
The recovered signal peaks at $2.099\,{\rm Hz}$ with a FWHM of $0.033\,{\rm Hz}$. 
There are also two visible harmonic peaks at integer multiples of the main peak. 
Their amplitudes as a fraction of the height of the main peak are  $0.103$ and $0.098$ for the $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks respectively. 
}

Continuous-wave searches look for signals which are close to monochromatic. 
In this section we consider a simple sinusoidal tone; a note at a single constant frequency. 
As described in Section~\ref{sec:ifo}, the audio is played through a speaker fixed to the back of one of the mirrors.

@@

We take data from a single off-centre pixel indicated by the red dot in Fig.~\ref{fig:interference_pattern}. 
The green-channel of the video is used as an approximation to the total intensity as the laser produces monochromatic green light (at $532\,{\rm nm}$).
As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording.
We measure a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. 
Two harmonics can also be seen in Fig.~\ref{fig:webcam_spectrum} at integer harmonics of the main peak. 
The $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks have amplitudes of $0.103$ and $0.098$ respectively as a fraction of the height of the main peak. 
They are probably due to the nonlinear system response discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}.

### 2020-08-16 paper/ifo-tableTopGWs.tex
small tidying in the text

\begin{figure}
% 	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam.pdf}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. 
The beam is first reflected from a mirror at a right angle and then focused at a converging lens before the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The beam is enlarged by a diverging lens after the beamsplitter. 
The interference pattern is projected onto a screen and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 provides the signal input from the PC.  
%Michelson interferometer schematic with interference pattern recorded by a commercial webcam. The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. The beam is first reflected from a mirror a right angle (to save space on the optical breadboard) and then focused at a converging lens before the $50$/$50$ beamsplitter. The split beam is reflected from mirrors M1 and M2 at the ends of the interferometer arms and recombined at the beamsplitter. The interference pattern is enlarged by a diverging lens after the beamsplitter and projected onto a screen. A commertial speaker is fixed to the back of M2, as the signal input. The interferometer arms lengths are roughly $3\,{\rm cm}$ different and the entire interferometer (without the screen) fits within the dimensions of an A4 piece of paper. We did not include a compensating plate with the beamsplitter, as is often done.} 
    }
\end{figure}

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=.45\textwidth]{figures/webcam_still0_crop.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
Michelson interference pattern as viewed by an off-axis webcam (which is why the pattern is elliptical). 
The red dot marks where the intensity time series is measured. 
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
% Interference pattern from Michelson interferometer as viewed by webcam from off-centre (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured from. For scale, the central fringe is approximately $5\,{\rm mm}$ across.
}
\end{figure}

The interference pattern is a ring of interference fringes as shown by the photograph in Fig.~\ref{fig:interference_pattern}. 
The pattern is enlarged by a diverging lens between the beamsplitter and the screen to aid in viewing. 
A change in the relative length of the interferometer arms results in the interference rings moving radially inwards or outwards. 
Ring interference patterns like this are often used in gravitational-wave table-top demonstrations to allow changes in the pattern to be easily viewed. 
%This results in a phase difference between the two beams when they meet at the screen. This path difference is perhaps better seen as the perpendicular separation between one mirror (M1) and the virtual image (M2\textquotesingle) of the other. For an aligned configuration and a fixed separation, the phase difference at a point on the screen only depends on the angle from the centre-line of the beam, and so the pattern is circularly symmetric. When the path difference changes, the fringes stay circular but move radially in or out.
%A ring interference pattern is often used in gravitational-wave table-top demonstrations to allow changes in the pattern to be easily seen.

A $0.5\,{\rm W}$ speaker is fixed to the back of one of the mirrors (M2 in Fig.~\ref{fig:ifo_schematic_webcam}), providing a controllable source of vibrations. 
The speaker is one half of a pair of commercial, USB-powered speakers fed by a $3.5\,{\rm mm}$ jack and driven by a PC. 
It is attached to the back of the mirror using commercial adhesive putty. 
The other speaker in the pair is kept face-down and as far away from the interferometer as the cables to reduce its impact as a second source of vibrations.

%As the speaker deforms to play sound, the mirror moves producing a change in arm length related to some coupling function of the amplitude and frequency of the audio. 
The motion of the interference fringes follows the motion of the mirror, and therefore the speaker. 
If the fringe motion is small, then changes in the intensity at some point on the screen occur at the same frequency as the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.
Even without over-counting, large fringe motions introduce a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome - but physically interesting! - phenomena like frequency doubling.
We derive the non-linear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photo-diode. 
In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
In Section~\ref{sec:optical_microphone}, which considers more complex audio, we turn to a photodiode because it offers a higher sample rate than the webcam.

\begin{comment}
Even though gravitational waves are caused by some of the most massive events in the universe, their observable effects on Earth are vanishingly small | on the order of only one part in $10^{21}$ for our best detectors~\cite{GW150914}. Clearly, the measurement of such effects is far beyond any table-top experiment. But, the fundamental principle of gravitational wave detection remains that changes in the lengths of the beam arms of an interferometer leads to a change in the interference pattern observed. By observing sound waves that vibrate the mirrors of a table-top interferometer and so change the lengths of the beam arms we have an analogue of this principle.

\subsection{Method of Michelson interferometry to observe sound}
% hardware: path difference, virtual thin film interference

In a Michelson interferometer, a coherent (monochromatic and equal phase) source of laser light is split evenly by a beamsplitter into two arms at right-angles to each other, travelling down and back from a mirror at the end of each arm. The beams recombine and interfere at the beamsplitter before hitting a screen to produce a pattern of concentric, circular fringes. The configuration used in this experiment is shown diagrammatically in Figure~\ref{fig:ifo_schematic_webcam} and as part of a photo in Figure~\ref{fig:setup_pic2}.

A sample interference pattern is shown in Figure~\ref{fig:interference_pattern}.
This pattern is due to the path difference between the two arms of unequal lengths (if the lengths were equal, then only a single, bright dot would be seen). This results in a phase difference between the two beams when they meet at the screen. This path difference is perhaps better seen as the perpendicular separation between one mirror (M1) and the virtual image (M2\textquotesingle) of the other. For an aligned configuration and a fixed separation, the phase difference at a point on the screen only depends on the angle from the centre-line of the beam, and so the pattern is circularly symmetric. When the path difference changes, the fringes stay circular but move radially in or out.

We used a 0.5W speaker stuck to the back of one of the mirrors as an audio source for our demonstration. This speaker was part of a commercial 3.5mm-jack, USB powered pair of speakers and was driven by a PC. To attach the speaker to one of the mirrors, we wedged it into part of the mirror mount and used commercial adhesive putty to keep it in place. We kept the other speaker in the pair face-down and as far away from the experiment as the cables would allow to reduce it acting as a second source.

As the speaker deforms to play sound, the mirror moves with a change in separation as some coupling function of the amplitude and frequency of the sound. The motion of the fringes follows this motion of the mirror and therefore the speaker at the same frequency.
If the fringe motion is small, then changes in the intensity at some point on the screen follow the frequency of the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion was large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting would artificially raise the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.

For later sections, we derive the non-linear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen, this can be found in Appendix~\ref{app:intensity_derivation}.

% Note that we do not include the common compensating plate that equalises the amount of time spent in glass for each beam. We did not find it necessary to include it to be able to demonstrate the results desired.

\begin{figure}[h]
	\includegraphics[width=\textwidth]{figures/webcam_still0.pdf}
	\caption{Interference pattern from Michelson interferometer as viewed by webcam from off-centre (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured from. For scale, the central fringe is approximately 5mm across}
	\label{fig:interference_pattern}
\end{figure}
\end{comment}

>>>>

\begin{figure}
	\includegraphics[width=0.49\textwidth]{figures/ifo_schematic_webcam_edit.pdf}
	\caption{\label{fig:ifo_schematic_webcam}
Schematic of the table-top Michelson interferometer. 
The laser (top-left) is a Class 2 $532\,{\rm nm}$ laser~\cite{ThorLabsIFO}. 
The beam is first reflected from a mirror at a right angle and then focused at a converging lens before reaching the beamsplitter which is $50$/$50$ in reflection and transmission. 
The split beams are reflected from mirrors (M1 and M2) at the ends of the interferometer arms and are recombined at the beamsplitter. 
The beam is enlarged by a diverging lens after the beamsplitter. 
The interference pattern is projected onto a screen and recorded using a commercial webcam connected to a personal computer (PC). 
A speaker fixed to the back of M2 provides the signal input from the PC.  
    }
\end{figure}

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=.45\textwidth]{figures/webcam_still0_crop.pdf}
 \end{center}
 \caption{\label{fig:interference_pattern}
Michelson interference pattern as viewed by an off-axis webcam (which is why the pattern is elliptical). 
The red dot marks where the intensity time series is measured. 
The central bright fringe of the interference pattern is approximately $5\,{\rm mm}$ in diameter. 
}
\end{figure}

The interference pattern is a ring of interference fringes as shown by the photograph in Fig.~\ref{fig:interference_pattern}. 
The pattern is enlarged by a diverging lens between the beamsplitter and the screen to aid in viewing. 
A change in the relative length of the interferometer arms results in the interference rings moving radially inwards or outwards. 
Ring interference patterns like this are often used in gravitational-wave table-top demonstrations to allow changes in the pattern to be easily viewed.

A $0.5\,{\rm W}$ speaker is fixed to the back of one of the mirrors (M2 in Fig.~\ref{fig:ifo_schematic_webcam}), providing a controllable source of vibrations. 
The speaker is one half of a pair of commercial, USB-powered speakers fed by a $3.5\,{\rm mm}$ jack and driven by a PC. 
It is attached to the back of the mirror using commercial adhesive putty. 
The other speaker in the pair is kept face-down and as far away from the interferometer as the cables allow to reduce its impact as a second source of vibrations.

The motion of the interference fringes follows the motion of the mirror, and therefore the speaker. 
If the fringe motion is small, then changes in the intensity at some point on the screen occur at the same frequency as the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion is large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting artificially raises the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.
Even without over-counting, large fringe motions introduce a nonlinear relation between the intensity fluctuations and injected audio, leading to troublesome---but physically interesting!---phenomena like frequency doubling.
We derive the nonlinear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen in Appendix~\ref{app:intensity_derivation}.

Data can be recorded via a webcam (as shown in Fig.~\ref{fig:ifo_schematic_webcam}) or photodiode. 
In Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering} a webcam is used. 
In Section~\ref{sec:optical_microphone}, which considers more complex audio, we turn to a photodiode because it offers a higher sample rate than the webcam.

### 2020-08-16 paper/ifo-wanderingTone.tex
small tidying in the text

One method to search for a wandering signal is to split the data into several shorter segments which are analysed individually. 
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies with as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}). 
In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find the most likely wandering frequency signal through the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
It is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

\subsection{The Viterbi algorithm}
\label{sec:viterbi}

@@

%Here we overview the Viterbi algorithm and describe its usage in this work. 
%Thinking about this 
A spectrogram is a grid indexed by time $t$ and frequency $f$ where here there are $N_t$ and $N_f$ time and frequency bins respectively, as shown in Fig.~\ref{fig:viterbi}.
We label the elements of the grid as $t_i$ and $f_j$ with $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. 
The value in each grid element is a detection statistic based on the time series data observed. 
It indicates the likelihood that a signal is present in the element. 
We label the normalised Fourier amplitude as $F(t_i,f_j)$. 
%In continuous wave searches the detection statistic depends on the type of gravitational wave source. 
%In this work we use the normalised Fourier amplitude, which we label $F(t_i,f_j)$.

The objective is to find the most likely path through the grid given the data and any probabilistic constraints on how the frequency wanders from $t_i$ to $t_{i+1}$. 
In continuous wave searches, a physical model is used to decide how much the frequency of the signal is allowed to wander at each time step, which depends on the type of target. 
In this work, at each time step we allow the frequency to either a) stay in the same bin; b) move up a single frequency bin; or c) move down a single frequency bin. 
The probability of transitioning from $f_j$ at $t_i$ to any of $\{f_{j+1},f_j,f_{j-1}\}$ at $t_{i+1}$ is called the transition probability $A(f_k,f_m)$. 
In this work the three options have equal probability $A(f_k,f_m)=1/3$ for $k=m+1,m,m-1$ and $A(f_k,f_m)=0$ otherwise.

A path through the grid has a likelihood dependent on the detection statistic (the observation) and the transition probability.
If the starting probabilities are all equal, i.e. $Pr[f(t_0)] = 1/N_f$, the the probability of the frequency equalling $f(t_{i+1})$ at $t = t_{i+1}$ is defined recursively by
%If there is equal probability of starting $f_j$ at $t=t_0$, the probability of path $Q$ given the observation $O$ is 
\begin{eqnarray}
%P(Q|O) =& A(f_j) F(t=N_t,f_j), \nonumber \\ &~~ \times \dots, A(f_j) F (t=t_0,f_j).
Pr[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &*A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     & *Pr[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes $Pr[f(t_{N_t}) = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

@@

\item At $t=t_{N_t}$, each $f_j$ has a value $Pr[f(t_{N_t})=f_j]$ (equation~\ref{eqn:recursiveViterbi}). 
The algorithm selects the terminating frequency $f(t_{N_t})$ with the highest value, marked (b) in Fig.~\ref{fig:viterbi}.

>>>>

One method to search for a wandering signal is to split the data into several shorter segments which are analysed individually. 
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}. 
In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin.

In continuous wave searches a hidden Markov model is used to describe the system. 
The frequency (the state) of the signal is unknown (hidden) and the signal is Markovian in that its current state depends only on the previous state. 
The hidden Markov model method relates the observable quantity (the detection statistic) to the signal frequency (the hidden state). 
The task is then to find the most likely wandering frequency signal given the observable of the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
This method is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

\subsection{The hidden Markov model and Viterbi algorithm}
\label{sec:viterbi}

@@

The spectrogram grid is indexed by time $t$ and frequency $f$ where there are $N_t$ and $N_f$ time and frequency bins respectively, as shown in Fig.~\ref{fig:viterbi}.
We label the elements of the grid as $t_i$ and $f_j$ with $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. 
The normalised Fourier amplitude $F(t_i,f_j)$ is the detection statistic indicating the likelihood that a signal is present in each grid element.

The objective is to find the most likely path through the grid given the data and any probabilistic constraints on how the frequency wanders from $t_i$ to $t_{i+1}$. 
In continuous wave searches, a physical model is used to decide how much the frequency of the signal is allowed to wander at each time step, which depends on the type of target. 
In this work, at each time step we allow the frequency to either a) stay in the same bin; b) move up a single frequency bin; or c) move down a single frequency bin. 
In a hidden Markov model, the probability of transitioning from $f_j$ at $t_i$ to any of $\{f_{j+1},f_j,f_{j-1}\}$ at $t_{i+1}$ is called the transition probability $A(f_k,f_m)$. 
In this work the three options have equal probability $A(f_k,f_m)=1/3$ for $k=m+1,m,m-1$ and $A(f_k,f_m)=0$ otherwise.

A path through the grid has a likelihood dependent on the detection statistic (the observation) and the transition probability.
If the starting probabilities are all equal, i.e. ${\rm Pr}[f(t_0)] = 1/N_f$, the the probability of the frequency equalling $f(t_{i+1})$ at $t = t_{i+1}$ is defined recursively by
\begin{eqnarray}
{\rm Pr}[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &\times A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     &\times {\rm Pr}[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes ${\rm Pr}[f(t_{N_t}) = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

@@

\item At $t=t_{N_t}$, each $f_j$ has a value ${\rm Pr}[f(t_{N_t})=f_j]$ (equation~\ref{eqn:recursiveViterbi}). 
The algorithm selects the terminating frequency $f(t_{N_t})$ with the highest value, marked (b) in Fig.~\ref{fig:viterbi}.
